LCS Tool : A Computational Platform for Lagrangian Coherent Structures

### 2014-09-05 main.tex
Replace explicit reference numbers by LaTeX label and reference tags

\lstloadlanguages{Matlab}
\lstset{basicstyle=\small,breaklines=true,frame=single,float,columns=fullflexible,numbers=left,stepnumber=2,backgroundcolor=\color{gray}}

@@

\begin{table}
\begin{enumerate}
\item Define a local maximization distance.
\item Find all points of the main grid that are local maxima of $\lambda_2$ within a circle whose radius is the local  maximization distance.
\item Define a maximum shrinkline length
\item Integrate a shrinkline forward and backward according to \cref{eq:strainline} and using the largest $\lambda_2$ local maximum as the initial position. Integrate until the shrinkline has attained the maximum shrinkline length, or until it has reached the domain boundary.
\item Flag any remaining local maxima of $\lambda_2$ within the maximization distance of the shrinkline as ineligible initial positions for subsequent shrinklines
\item Continue integrating shrinklines using local maxima of $\lambda_2$ as initial positions until no eligible local maxima of $\lambda_2$ remain.
\item Remove all shrinkline segments within elliptic LCSs.
\end{enumerate}
\caption{Algorithm to calculate initial positions of repelling LCSs at time $t_0$. The algorithm for attracting LCSs is similar.}
\label{t:Hyperbolic LCS algorithm}
\end{table}

@@

As usual, the local maximization distance is set larger for stretchlines than for shrinklines (cf. l.\,4 and l.\,9). The purpose of the maximization distance is to obtain spatially separated LCSs and to avoid a dense tangle of lines that basically indicates the same hyperbolic LCS (cf. \cref{t:Hyperbolic LCS algorithm}, point 5). Setting the local maximization distance for stretchlines larger than for shrinklines allows obtaining a comparable number of stretchlines and shrinklines overall in the flow domain (recall that hyperbolic LCS seed points are discarded if they are within the local maximization distance of an existing hyperbolic LCS). Shrinklines are locally tangent to ridges of $\lambda_2$ maxima, whereas stretchlines are locally normal to these ridges. Setting the local maximization distance of stretchlines and shrinklines equal would therefore produce a greater number of stretchlines than shrinklines overall.

\lstinputlisting[caption={LCS Tool commands for double gyre hyperbolic LCSs. Condensed version of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m! },label=lst:double_gyre_lambda_strain_stretch_lcs]{listings/double_gyre/lambda_strain_stretch_lcs.m}

@@

The data is given at a spatial resolution of $1/4°$ and a temporal resolution of $7\,\mathrm{days}$. Due to the discrete data, defining the right hand side of \cref{eq:dtlon} and \cref{eq:dtlat} involves spline interpolation in space and time. An interpolant is generated first, then the function \lstinline!flowdata_derivative! evaluates the interpolants for the zonal and meridional velocity at the needed coordinates. \Cref{l:griddedInterp} shows the relevant part of the code in LCS Tool's ocean demo file \lstinline!hyperbolic_shear_lcs.m!. In lines 8-11, the commands for the interpolation are given of the velocity dataset are given.

\begin{lstlisting}[caption={LCS ocean demo: Parts of the demo script included in LCS Tool to compute LCS from an ocean dataset},label=l:griddedInterp]
% Input parameters
domain = [0,6;-34,-28];
resolution = [400,400];
timespan = [100,130];

% Velocity definition
...
interpMethod = 'spline';
vlon_interpolant = griddedInterpolant({time,lat,lon},vlon,interpMethod);
vlat_interpolant = griddedInterpolant({time,lat,lon},vlat,interpMethod);
lDerivative = @(t,x,~)flowdata_derivative(t,x,vlon_interpolant,vlat_interpolant);
incompressible = true;

% LCS parameters
% Cauchy-Green strain
cgEigenvalueFromMainGrid = false;
cgAuxGridRelDelta = 0.01;

@@

% Stretchlines
stretchlineMaxLength = 20;
stretchlineLocalMaxDistance = 4*gridSpace;
stretchlineOdeSolverOptions = odeset('relTol',1e-4);
...
% Cauchy-Green strain eigenvalues and eigenvectors
[cgEigenvector,cgEigenvalue] = eig_cgStrain(lDerivative,domain,resolution,timespan,'incompressible',incompressible,'eigenvalueFromMainGrid',cgEigenvalueFromMainGrid,'auxGridRelDelta',cgAuxGridRelDelta);

% Lambda-line LCSs
% Define Poincare sections; ...
poincareSection = struct('endPosition',{},'numPoints',{},'orbitMaxLength',{});
...
poincareSection(1).endPosition = [3.3,-32.1;3.7,-31.6];
poincareSection(2).endPosition = [1.3,-30.9;1.9,-31.1];

@@

% Set maximum orbit length to twice the expected circumference
nPoincareSection = numel(poincareSection);
for i = 1:nPoincareSection
    rOrbit = hypot(diff(poincareSection(i).endPosition(:,1)),diff(poincareSection(i).endPosition(:,2)));
    poincareSection(i).orbitMaxLength = 2*(2*pi*rOrbit);
end
[shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda);
closedLambdaLine = poincare_closed_orbit_multi(domain,resolution,shearline,poincareSection,'odeSolverOptions',lambdaLineOdeSolverOptions);
...
% Hyperbolic strainline LCSs
strainlineLcs = seed_curves_from_lambda_max(strainlineLocalMaxDistance,strainlineMaxLength,cgEigenvalue(:,2),cgEigenvector(:,1:2),domain,resolution,'odeSolverOptions',strainlineOdeSolverOptions);
...
stretchlineLcs = seed_curves_from_lambda_max(stretchlineLocalMaxDistance,stretchlineMaxLength,-cgEigenvalue(:,1),cgEigenvector(:,3:4),domain,resolution,'odeSolverOptions',stretchlineOdeSolverOptions);
\end{lstlisting}

We choose the integration time as $T=30$ days (\cref{l:griddedInterp}, line 4), which is larger than the eddy turnover time in this region. The resolution of the main computational grid for initial conditions is set to $400 \times 400$ (line 3). This corresponds to a resolution of roughly 0.015° and gives good results. With this choice, the resolution of the tracer grid is 15 times higher than the resolution of the velocity field. The flow is integrated and the Cauchy-Green strain tensor is computed by the function \lstinline!eig_cgStrain! (line 35). Incompressibility of the flow is enforced (line 12), the auxiliary grid distance is set to 1\% of the main grid distance (line 17), and eigenvalues are computed from the auxiliary grid (line 16). Elliptic LCSs are computed in line 54, after the Poincare sections have been set (line 41-42) and the $\eta_\pm^\lambda$ fields have been defined (line 53). Hyperbolic LCSs are computed in lines 57 and 59.

>>>>

\lstloadlanguages{Matlab}
\lstset{
  basicstyle=\small,
  breaklines=true,
  frame=single,
  float,
  columns=fullflexible,
  numbers=left,
  stepnumber=2,
  backgroundcolor=\color{gray},
  escapeinside={\%@}{\^^M}
}

@@

\begin{table}
\begin{enumerate}
\item Define a local maximization distance.
\item Find all points of the main grid that are local maxima of $\lambda_2$ within a circle whose radius is the local  maximization distance.
\item Define a maximum shrinkline length
\item Integrate a shrinkline forward and backward according to \cref{eq:strainline} and using the largest $\lambda_2$ local maximum as the initial position. Integrate until the shrinkline has attained the maximum shrinkline length, or until it has reached the domain boundary.
\item Flag any remaining local maxima of $\lambda_2$ within the maximization distance of the shrinkline as ineligible initial positions for subsequent shrinklines
\label{i:flag maxima}
\item Continue integrating shrinklines using local maxima of $\lambda_2$ as initial positions until no eligible local maxima of $\lambda_2$ remain.
\item Remove all shrinkline segments within elliptic LCSs.
\end{enumerate}
\caption{Algorithm to calculate initial positions of repelling LCSs at time $t_0$. The algorithm for attracting LCSs is similar.}
\label{t:Hyperbolic LCS algorithm}
\end{table}

@@

As usual, the local maximization distance is set larger for stretchlines than for shrinklines (cf. \cref{ll:strainlineLocalMaxDistance} and \cref{ll:stretchlineLocalMaxDistance}). The purpose of the maximization distance is to obtain spatially separated LCSs and to avoid a dense tangle of lines that basically indicates the same hyperbolic LCS (cf. \cref{t:Hyperbolic LCS algorithm}, \cref{i:flag maxima}). Setting the local maximization distance for stretchlines larger than for shrinklines allows obtaining a comparable number of stretchlines and shrinklines overall in the flow domain (recall that hyperbolic LCS seed points are discarded if they are within the local maximization distance of an existing hyperbolic LCS). Shrinklines are locally tangent to ridges of $\lambda_2$ maxima, whereas stretchlines are locally normal to these ridges. Setting the local maximization distance of stretchlines and shrinklines equal would therefore produce a greater number of stretchlines than shrinklines overall.

\lstinputlisting[
caption={LCS Tool commands for double gyre hyperbolic LCSs. Condensed version of the LCS Tool script \lstinline!demo/double_gyre/hyperbolic_shear_lcs.m!},
label=lst:double_gyre_lambda_strain_stretch_lcs
]
{listings/double_gyre/lambda_strain_stretch_lcs.m}

@@

The data is given at a spatial resolution of $1/4°$ and a temporal resolution of $7\,\mathrm{days}$. Due to the discrete data, defining the right hand side of \cref{eq:dtlon} and \cref{eq:dtlat} involves spline interpolation in space and time. An interpolant is generated first, then the function \lstinline!flowdata_derivative! evaluates the interpolants for the zonal and meridional velocity at the needed coordinates. \Cref{l:griddedInterp} shows the relevant part of the code in LCS Tool's ocean demo file \lstinline!hyperbolic_shear_lcs.m!. In \crefrange{ll:interpolation start}{ll:interpolation end}, the commands for the interpolation are given of the velocity dataset are given.

\begin{lstlisting}[caption={LCS ocean demo: Parts of the demo script included in LCS Tool to compute LCS from an ocean dataset},label=l:griddedInterp]
% Input parameters
domain = [0,6;-34,-28];
resolution = [400,400]; %@\label{ll:resolution}
timespan = [100,130]; %@\label{ll:timespan}

% Velocity definition
...
interpMethod = 'spline'; %@\label{ll:interpolation start}
vlon_interpolant = griddedInterpolant({time,lat,lon},vlon,interpMethod);
vlat_interpolant = griddedInterpolant({time,lat,lon},vlat,interpMethod);
lDerivative = @(t,x,~)flowdata_derivative(t,x,vlon_interpolant,vlat_interpolant); %@\label{ll:interpolation end}
incompressible = true; %@\label{ll:incompressible}

% LCS parameters
% Cauchy-Green strain
cgEigenvalueFromMainGrid = false; %@\label{ll:cgEigenvalueFromMainGrid}
cgAuxGridRelDelta = 0.01; %@\label{ll:cgAuxGridRelDelta}

@@

% Stretchlines
stretchlineMaxLength = 20;
stretchlineLocalMaxDistance = 4*gridSpace;
stretchlineOdeSolverOptions = odeset('relTol',1e-4);
...
% Cauchy-Green strain eigenvalues and eigenvectors
[cgEigenvector,cgEigenvalue] = eig_cgStrain(lDerivative,domain,resolution,timespan,'incompressible',incompressible,'eigenvalueFromMainGrid',cgEigenvalueFromMainGrid,'auxGridRelDelta',cgAuxGridRelDelta); %@\label{ll:eig_cgStrain}

% Lambda-line LCSs
% Define Poincare sections; ...
poincareSection = struct('endPosition',{},'numPoints',{},'orbitMaxLength',{});
...
poincareSection(1).endPosition = [3.3,-32.1;3.7,-31.6]; %@\label{ll:poincareSection(1)}
poincareSection(2).endPosition = [1.3,-30.9;1.9,-31.1]; %@\label{ll:poincareSection(2)}

@@

% Set maximum orbit length to twice the expected circumference
nPoincareSection = numel(poincareSection);
for i = 1:nPoincareSection
    rOrbit = hypot(diff(poincareSection(i).endPosition(:,1)),diff(poincareSection(i).endPosition(:,2)));
    poincareSection(i).orbitMaxLength = 2*(2*pi*rOrbit);
end
[shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda); %@\label{ll:lambda_line}
closedLambdaLine = poincare_closed_orbit_multi(domain,resolution,shearline,poincareSection,'odeSolverOptions',lambdaLineOdeSolverOptions); %@\label{ll:poincare_closed_orbit_multi}
...
% Hyperbolic strainline LCSs
strainlineLcs = seed_curves_from_lambda_max(strainlineLocalMaxDistance,strainlineMaxLength,cgEigenvalue(:,2),cgEigenvector(:,1:2),domain,resolution,'odeSolverOptions',strainlineOdeSolverOptions); %@\label{ll:strainlineLcs}
...
stretchlineLcs = seed_curves_from_lambda_max(stretchlineLocalMaxDistance,stretchlineMaxLength,-cgEigenvalue(:,1),cgEigenvector(:,3:4),domain,resolution,'odeSolverOptions',stretchlineOdeSolverOptions); %@\label{ll:stretchlineLcs}
\end{lstlisting}

We choose the integration time as $T=30$ days (\cref{l:griddedInterp}, \cref{ll:timespan}), which is larger than the eddy turnover time in this region. The resolution of the main computational grid for initial conditions is set to $400 \times 400$ (\cref{ll:resolution}). This corresponds to a resolution of roughly 0.015° and gives good results. With this choice, the resolution of the tracer grid is 15 times higher than the resolution of the velocity field. The flow is integrated and the Cauchy-Green strain tensor is computed by the function \lstinline!eig_cgStrain! (\cref{ll:eig_cgStrain}). Incompressibility of the flow is enforced (\cref{ll:incompressible}), the auxiliary grid distance is set to 1\% of the main grid distance (\cref{ll:cgAuxGridRelDelta}), and eigenvalues are computed from the auxiliary grid (\cref{ll:cgEigenvalueFromMainGrid}). Elliptic LCSs are computed in \cref{ll:poincare_closed_orbit_multi}, after the Poincare sections have been set (\cref{ll:poincareSection(1),ll:poincareSection(2)}) and the $\eta_\pm^\lambda$ fields have been defined (\cref{ll:lambda_line}). Hyperbolic LCSs are computed in \cref{ll:strainlineLcs,ll:stretchlineLcs}.
